Reviving the Lost Art of Combinator Graph Reduction

### 2023-01-23 experiments/superg/doc/paper/paper.tex
Write Kiselyov's algorithms comparison.

@@

\begin{figure*}
    \includesvg[width=\textwidth]{turner_vs_miranda}
    \centering
    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
    \label{fig:turner-vs-miranda}
\end{figure*}

@@

\begin{figure*}
    \includesvg[width=\textwidth]{turner_vs_tigre}
    \centering
    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
    \label{fig:turner-vs-tigre}
\end{figure*}

@@

\subsection{Kiselyov's algorithms}

\textbf{TEXT BELOW ARE NOTES!}
Programs:
\begin{itemize}
    \item ackermann
    \item digits of e
    \item linfib
    \item lsort
    \item nfib
    \item primes
    \item queens
    \item tak
    \item towers of hanoi
    \item treesort
\end{itemize}

Evaluate performance of:

\begin{itemize}
    \item Bracket abstraction
    \item Kiselyov
    \item GHC
\end{itemize}

Suggestion from \textit{Common Benchmarking Pitfalls}: Always look at your distribution.
When you have run benchmarks, take a look at the times for each sample.

@@

>>>>

\usepackage{pgfplots}

@@

\begin{figure*}
    \includesvg[width=\textwidth]{turner_vs_miranda}
    \centering
    \caption{
        Time to execute the Fibonacci benchmark program on the original Miranda and our Miranda-style engine.
        A lower value on the Y-axis is better.}
    \label{fig:turner-vs-miranda}
\end{figure*}

@@

\begin{figure*}
    \includesvg[width=\textwidth]{turner_vs_tigre}
    \centering
    \caption{
        Time to execute the Fibonacci benchmark program on Miranda-style and TIGRE-style engines.
        A lower value on the Y-axis is better.}
    \label{fig:turner-vs-tigre}
\end{figure*}

@@

\subsection{Kiselyov's algorithms}
To compare the semantic translation algorithms from~\cite{kiselyov_lambda_2018} with each other, we again use the Fibonacci test program and execute all compiled programs on the Miranda-style engine.
The bracket abstraction algorithm has been added as a reference point.
The results can be seen in figure~\ref{fig:bracket-vs-kiselyov}.

\begin{figure*}
    \includesvg[width=\textwidth]{bracket_vs_kiselyov}
    \centering
    \caption{Time to execute the Fibonacci benchmark program compiled with the algorithms from~\cite{kiselyov_lambda_2018}. A lower value on the Y-axis is better.}
    \label{fig:bracket-vs-kiselyov}
\end{figure*}

All of the algorithms presented in~\cite{kiselyov_lambda_2018} produce more efficient code than the reference bracket abstraction algorithm.
The $\lazyeta$ compilation scheme produces the best code for the Fibonacci program.
For many of the test programs in~\cite{kiselyov_lambda_2018} it was shown to generate the shortest programs of all the algorithms presented there.
This is also true for the Fibonacci program, as shown in figure~\ref{fig:fib-sizes}.

\begin{figure}
    \begin{tikzpicture}
        \begin{axis}[
                xlabel={Compiled expression size (nodes)},
                xmin=0,
                ylabel={Translation algorithm},
                symbolic y coords={Bracket,$\strict$,$\lazy$,$\lazyeta$,$\linear$},
            ]
            \addplot[xbar] coordinates {
                    (73,Bracket)
                    (59,$\strict$)
                    (59,$\lazy$)
                    (47,$\lazyeta$)
                    (53,$\linear$)
                };
        \end{axis}
    \end{tikzpicture}
    \centering
    \caption{Size of combinator expression generated for the Fibonacci program by different translation algorithm}
    \label{fig:fib-sizes}
\end{figure}

It comes as no surprise that the shortest translation is also the fastest to execution, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
We see a similar pattern for the Ackermann function, for which results may be found in appendix~\ref{app:ack}.
This suggests that while the semantic translation algorithms from~\cite{kiselyov_lambda_2018} reduce the size of the combinator expression, they reduce runtime even further.

\section{Related work}
supercombinators

GHC

Kiselyov's later work.

Edward Kmett's SIMD combinator stuff.

@@

\newpage

\subsection{Ackermann benchmark results}
\label{app:ack}

\begin{figure*}
    \begin{lstlisting}
(defun ack (x z) (if (= x 0)
                     (+ z 1)
                     (if (= z 0)
                         (ack (- x 1) 1)
                         (ack (- x 1) (ack x (- z 1))))))
    \end{lstlisting}
    \centering
    \caption{
        Ackermann's function.
    }
    \label{fig:ack}
\end{figure*}

\begin{figure*}
    \includesvg[width=\textwidth]{ack_bracket_vs_kiselyov}
    \centering
    \caption{Time to execute the Ackermann benchmark program compiled with the algorithms from~\cite{kiselyov_lambda_2018}. A lower value on the Y-axis is better.}
    \label{fig:ack-bracket-vs-kiselyov}
\end{figure*}

\begin{figure}
    \begin{tikzpicture}
        \begin{axis}[
                xlabel={Compiled expression size (nodes)},
                xmin=0,
                ylabel={Translation algorithm},
                symbolic y coords={Bracket,$\strict$,$\lazy$,$\lazyeta$,$\linear$},
            ]
            \addplot[xbar] coordinates {
                    (415,Bracket)
                    (175,$\strict$)
                    (137,$\lazy$)
                    (109,$\lazyeta$)
                    (123,$\linear$)
                };
        \end{axis}
    \end{tikzpicture}
    \centering
    \caption{Size of combinator expression generated for the Ackermann program by different translation algorithm}
    \label{fig:ack-sizes}
\end{figure}
